miniKanren with fair search strategies

### 2019-04-01 Paper/paper.tex
revise intro, biDFS.

\subsection{fairness}

A disjunction is fair if when a corresponding goal is queried, answers of 
lower costs come first. The \emph{cost} of an answer the number of relation 
applications needed to find the answer. This idea is borrowed from Silvija 
Seres's work [*].

Interleaving DFS, the search strategy of TRS2 miniKanren, is not fair. Among 
the search strategies we devised, biDFS is not fair, but is more fair than 
iDFS. And BFS is fair.

\subsection{related works}

The unfairness of disjunction has been noticed by Seres et al. Their complete 
and fair search strategy is also named breadth-first search. Their BFS 
is fair and is similar to ours. However, their Haskell implementation cannot be 
translated to Scheme directly. This is partially due to the difference if 
calling convensions of host languages. Another reason is that their search 
space are infinite even when no answers exists. This is not feasible 
in miniKanren, where users can query all answers.

% TODO: summarize following sections) Section 2 explains why iDFS biases toward 
% left \texttt{conde} clauses. Section 3 is about the biDFS. Section 4 is 
% about BFS.

\section{representation of search space}

``Stream'' is often used to name the representation of search space in 
miniKanren. However, the search space is more like a stream of list, 
because the information that a stream has no \texttt{car} is employed in all 
complete miniKanren search strategies to redirect search effort. If the search 
space is a normal stream, the information should have been used in a more 
trivial way.

In the rest of this paper, we call the \texttt{car}s of a stream 
its \textit{mature part}, and the last \texttt{cdr} its \textit{immature part}.

\section{why disj is unfair}

The multiarity disjuction operator \texttt{disj} combine its sub-goals with 
\texttt{disj2}, right associatively. \texttt{disj2} is fair in the sence that 
it allocates computational effort evenly to its two sub-goals. Unfortunately, 
the way how \texttt{disj2} achieves fairness leads to the unfairness of 
\texttt{disj}. The core functionality of \texttt{disj2} is completed by 
\texttt{append-inf} (Fig.~ref{disj2-and-append-inf}). When both input streams 
are infinite, the resulting mature part contains only the mature part of 
\texttt{s-inf}. The whole \texttt{t-inf} goes to the resulting immature part. 
However, \texttt{t-inf} and \texttt{s-inf} are swapped in the delayed recursive 
call. Hence the search strategy spend computational effort evenly in two 
appended streams. As \texttt{disj} applys \texttt{disj2} right associatively, 
the left clauses are more frequently explored.

\begin{figure}
  \lstinputlisting{Figures/disj2.rkt}
  \caption{\texttt{disj2} and \texttt{append-inf}}
  \label{disj2-and-append-inf}
\end{figure}

@@

A \texttt{conde} expression looks like a binary tree, if we consider 
\texttt{disj2}s as internal nodes and \texttt{conde} clauses as a terminal 
nodes. In fact, each of them is one of the most unbalanced binary trees. The 
basic idea of our first solution, balanced interleaving DFS, is to make the 
tree balanced. We introduce a function \texttt{disj*} and its helper 
\texttt{split} (Fig.~\ref{disj*}). \texttt{disj*} essentially construct a 
balanced \texttt{disj2} tree. \texttt{split} splits elements of \texttt{ls} 
into two even half and pass them to the continuation parameter \texttt{k}.

\begin{figure}
  \lstinputlisting{Figures/disj*.rkt}
  \caption{\texttt{disj*}}
  \label{disj*}
\end{figure}

@@

\subsection{cost of answers}

%TODO the meaning of resolution step in Silvija Seres's paper

Now we illustrate the costs of answers by running a miniKanren relation.
Fig.~\ref{def-repeato} defines the relation \texttt{repeato} that 
relates a term \texttt{x} with a list whose elements are all \texttt{x}s.

The above \texttt{run} generates 4 answers. All are lists of \texttt{*}s.
The order of the answers reflects the order miniKanren discovers them:
the first answer in the list is first discovered. This result is not suprising:
to generate the first answer, \texttt{'()}, miniKanren needs to apply
\texttt{repeato} only once and the later answers need more recursive
applications. In this example, the cost of each answer is the same as
one more than the number of \texttt{*}s: the cost of \texttt{'()} is 1,
the cost of \texttt{'(*)} is 2, and so on.

A list of answer is in the \emph{cost-respecting} order if no answer occurs
before another answer of a lower cost. In the above example, the answers are
cost-respecting. The iDFS search, however, does not generate cost-respecting
answers in general. As an example, consider the following \texttt{run} of
\texttt{repeato}.
\begin{center}
  \begin{tabular}{c}
   \lstinputlisting{Figures/run-conde-repeato-iDFS.rkt}
   \end{tabular}
\end{center}

The results are not cost-respecting. For example, \texttt{'(a a)} occurs before
\texttt{'(b)} while \texttt{'(a a)} is associated with a higher cost. 
iDFS strategy is the cause, since it prioritizes the first \texttt{conde} case
considerablely. When every \texttt{conde} case are equally productive, the iDFS
strategy takes $1/2^{i}$ answers from the $i$-th case, except the last case,
which share the same portion as the second last one.

%TODO relation *application*

For the above \texttt{run}, both search strategies produces answers in 
increasing order of costs, i.e. both of them are \emph{cost-respecting}. In 
more complicated cases, however, interleaving DFS might not produce answers in 
cost-repecting order. For instance, with iDFS the \texttt{run} in Fig.~
\ref{conde-repeato-iDFS} produces answers in a seemingly random order. In 
contrast, the same run with BFS produces answers in an expected order (Fig.~
\ref{conde-repeato-BFS}).

\lstinputlisting{Figures/run-conde-repeato-BFS.rkt}

>>>>

\section{related works}

The unfairness of disjunction has been noticed by Seres et al. Their complete 
and fair search strategy is also named breadth-first search. Their BFS 
is fair and is similar to ours. However, their Haskell implementation cannot be 
translated to Scheme directly. This is partially due to the difference in 
calling convensions of host languages. Besides, their search space are infinite 
even when no answers exists. This is not feasible in miniKanren, where 
users can query all answers.

\section{fairness}

% KC: I changed the cost definition from ``... needed to *find* the answer''
%                                     to ``... needed to *verify* the answer''

A disjunction is fair if when a corresponding goal is queried, answers of 
lower costs come first. The \emph{cost} of an answer the number of relation 
applications needed to verify the answer. Now we illustrate the costs of answers 
by running a miniKanren relation. Fig.~\ref{def-repeato} defines the relation 
\texttt{repeato} that relates a term \texttt{x} with a list whose elements are 
all \texttt{x}s.

% COmmented out: This idea is borrowed from Silvija Seres's work [*].
% KC: Seres didn't phrase it this way.
%   We might need to compare our definition with Seres's one, instead of 
%   saying the idea is borrowed.

The above \texttt{run} generates 4 answers. All are lists of \texttt{*}s.
The order of the answers reflects the order miniKanren discovers them:
the first answer in the list is first discovered. This order is not suprising:
to generate the first answer, \texttt{'()}, miniKanren needs to apply
\texttt{repeato} only once and the later answers need more recursive
applications. In this example, the cost of each answer is the same as
one more than the number of \texttt{*}s: the cost of \texttt{'()} is 1,
the cost of \texttt{'(*)} is 2, and so on.

A query result (list of answers) is in the \emph{cost-respecting} order if no 
answer occurs before another answer of a lower cost. In the above example, the 
result is cost-respecting. The iDFS search, however, does not generate 
cost-respecting answers in general. As an example, consider the following 
\texttt{run} of \texttt{repeato}.

\begin{center}
  \begin{tabular}{c}
   \lstinputlisting{Figures/run-conde-repeato-iDFS.rkt}
   \end{tabular}
\end{center}

The results is not cost-respecting. For example, \texttt{'(a a)} occurs before
\texttt{'(b)} while \texttt{'(a a)} is associated with a higher cost. 
iDFS strategy is the cause, since it prioritizes the first \texttt{conde} case
considerablely. When every \texttt{conde} case are equally productive, the iDFS
strategy takes $1/2^{i}$ answers from the $i$-th case, except the last case,
which share the same portion as the second last one. In contrast, the same run 
with BFS produces answers in an expected order.

\begin{center}
  \begin{tabular}{c}
   \lstinputlisting{Figures/run-conde-repeato-BFS.rkt}
   \end{tabular}
\end{center}

% \section{representation of search space}
% 
% ``Stream'' is often used to name the representation of search space in 
% miniKanren. However, the search space is more like a stream of list, because the 
% information that a stream is thunk is employed to redirect search effort. If the 
% search space is a normal stream, the information should have been used in a more 
% trivial way. 
% 
% In the rest of this paper, we call the \texttt{car}s of a stream 
% its \textit{mature part}, and the last \texttt{cdr} its \textit{immature part}.
% 
% \section{why disj is unfair}
% 
% The multiarity disjuction operator \texttt{disj} combine its sub-goals with 
% \texttt{disj2}, right associatively. Interestingly, although \texttt{disj} is 
% unfair, \texttt{disj2} is fair. The core functionality of \texttt{disj2} is 
% completed by \texttt{append-inf} (Fig.~ref{disj2-and-append-inf}). When both 
% input streams are infinite, the resulting mature part contains only the mature 
% part of \texttt{s-inf}. The whole \texttt{t-inf} goes to the resulting immature 
% part. However, \texttt{t-inf} and \texttt{s-inf} are swapped in the delayed 
% recursive call. Hence the search strategy spend computational effort evenly in 
% two appended streams. As \texttt{disj} applys \texttt{disj2} right 
% associatively, the left clauses are more frequently explored.
% 
% \begin{figure}
%   \lstinputlisting{Figures/disj2.rkt}
%   \caption{\texttt{disj2} and \texttt{append-inf}}
%   \label{disj2-and-append-inf}
% \end{figure}

@@

Our first solution, balanced interleaving DFS, is based on the observation that 
a disjuction can be viewed as a binary tree, where \texttt{disj2}s are nodes 
and sub-goals are leaves. In iDFS, the tree is in one of the most unbalanced 
forms, because \texttt{disj} applys \texttt{disj2} right associatively. As 
\texttt{disj2} allocates computational effort interleavingly to its two 
sub-goals, a disjunction allocates half computational effort to its first 
sub-goal, quater to the second, eighth to the third, and so on.

The key idea of biDFS is to make disjunction trees balanced 
(Fig.~\ref{balanced-disj}). We introduce a function \texttt{disj*} and its 
helper \texttt{split}, and change the \texttt{disj} macro to call 
\texttt{disj*} immediately. \texttt{disj*} essentially construct a balanced 
\texttt{disj2} tree. The \texttt{split} helper splits elements of \texttt{ls} 
into two lists of roughly the same length, then apply \texttt{k} to them.

\begin{figure}
  \lstinputlisting{Figures/balanced-disj.rkt}
  \caption{\texttt{balanced-disj}}
  \label{balanced-disj}
\end{figure}

@@
